Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-15 paper.tex
Tweak intro

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is a very efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from a Wright-Fisher population, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data.
This efficiency, and the adaptability of the model to different
applications has led to dozens of different simulation tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.
Two technological developments of recent years, however,
pose major challenges to the majority of these methods.
% Some citations here? Darwin tree of life maybe?
Firstly, whole genome sequencing---and in particular fourth-generation
sequencing technologies---have made chromosome-level assemblies
commonplace. Thus, modelling genetic variation data as a series of unlinked loci,
a previously reasonable assumption, is no longer defensible and
we must fully account for recombination.
While a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more challenging,
and methods such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second major challenge to existing coalescent simulators is the
exponential increase in sample size of genetic studies. Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it is now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
introduced as part of \msprime\, makes it possible to store
such huge simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also led to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
This paper marks the release of \msprime\ 1.0.
We provide an overview of its extensive features,
demonstrate its performance advantages over alternative software,
and discuss opportunites for ongoing
open-source community-based development.

>>>>

The coalescent
process~\citep{kingman1982coalescent,hudson1983testing,tajima1983evolutionary}
models the ancestry of a set of sampled genomes,
providing a mathematical description of the
genealogical tree that relates the samples to one another.
It has proved to be a powerful model,
and is now central to
population genetics~\citep{hudson1990gene,hein2004gene,wakely2008coalescent}.
The coalescent is a very efficient framework for population genetic
simulation, because it allows us to simulate the genetic ancestry for
a sample from a Wright-Fisher population, without explicitly representing
the population in memory or stepping through the generations.
Indeed, \cite{hudson1983testing} independently derived the
% Anyone disagree here? That's the impression I get from the paper.
coalescent \emph{in order to} efficiently simulate data under a neutral model.
This efficiency, and the adaptability of the model to different
applications has led to dozens of different simulation tools
being developed over the decades~\citep{carvajal2008simulation,liu2008survey,
arenas2012simulation,yuan2012overview,hoban2012computer,yang2014critical,
peng2015genetic}.

Two technological developments of recent years
pose major challenges to the majority of existing simulation methods.
Firstly, fourth-generation sequencing technologies have made
complete chromosome-level assemblies possible~\citep{miga2020telomere},
% Some citations here? Darwin tree of life maybe?
and high quality assemblies are now available for many species.
Thus, modelling genetic variation data as a series of unlinked loci
is no longer a reasonable approximation, and
we must fully account for recombination.
While a genealogical tree relating $n$ samples in the single-locus coalescent
can be simulated in $O(n)$ time~\citep{hudson1990gene},
the coalescent with recombination is far more complex,
and programs such as
Hudson's classical \ms~\citep{hudson2002generating}
can only simulate short segments under the influence of recombination.
The second challenge facing simulation methods is that
sample sizes in genetic studies have grown very quickly
in recent years, enabled by the precipitous fall in genome sequencing costs.
Human datasets like the
UK Biobank~\citep{bycroft2018genome} and
gnomAD~\citep{karczewski2020mutational} now consist of hundreds of
thousands of genomes and many other datasets on a similar scale
are becoming available~\citep{tanjo2021practical}.
Classical simulators such as \ms\ and even fast approximate methods
such as \scrm~\citep{staab2015scrm} simply cannot
cope with such a large number of samples.

The \msprime\ simulator~\citep{kelleher2016efficient,kelleher2020coalescent}
has greatly increased the scope of coalescent simulations,
and it is now straightforward to simulate millions of whole chromosomes.
The ``succinct tree sequence'' data
structure~\citep{kelleher2016efficient,kelleher2018efficient,kelleher2019inferring,
wohns2021unified},
originally introduced as part of \msprime, makes it possible to store
such large simulations in a few gigabytes, several orders
of magnitude smaller than commonly used formats.
The succinct tree sequence has also led to major advances in forwards-time
simulation~\citep{kelleher2018efficient,haller2018tree},
ancestry inference~\citep{kelleher2019inferring,wohns2021unified}
and calculation of population genetic
statistics~\citep{kelleher2016efficient,ralph2020efficiently}.
Through a rigorous open-source community development process,
\msprime\ has gained a large number of features since its introduction,
making it a highly efficient and flexible platform for population
genetic simulation.
This paper marks the release of \msprime\ 1.0.
We provide an overview of its extensive features,
demonstrate its performance advantages over alternative software,
and discuss opportunites for ongoing
open-source community-based development.
